Bird beak's elaboration and exploration

### 2021-05-27 Analysis/03-elaboration_exploration_analyses.Rmd
loading data for elaboration part
Major axes equation

---
title: "03-elaboration exploration analyses"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Elaboration/exploration analyses

## Elaboration/exploration metrics

Measure the different elaboration vs exploration metrics with different vectors:

 - [ ] MRCA trajectory: the vector between the root of the tree and the OTU's descendant
 - [ ] MRCA average trajectory: the vector between the centroid of the group's nodes and the root of the tree
 - [ ] Evolutionary living average trajectory: the vector between the root of the tree and the centroid of the living taxa
 - [ ] Evolutionary overall average trajectory: the vector between the root of the tree and the centroid of the tips and nodes
 - [ ] Maximum living trajectory: the vector between the two most distant OTUs
 - [ ] Maximum evolutionary trajectory: the vector between the root and the OTU most distant from the root

## Elaboration/exploration in the present

Apply these metrics on both clades to get two scores for each taxa (elaboration vs. exploration plots).

## Elaboration/exploration through time

Do some time-slicing and run these metrics through time.

>>>>

---
title: "Bird beak's elaboration and exploration"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See <!-- §§§run_replicates_script --> for running the actual analyses.

# Elaboration and exploration analyses

```{r, eval = FALSE, echo = FALSE}
library(beer)
## Loading the MCMCglmm
load("../Data/processed/combined_results.rda")
## Loading the tree and data
data(demo_data)
data(demo_tree)
```

## Extracting the variance covariance matrices from the MCMCglmm mini-chains

To analyse the exploration and elaboration aspects of both clades and tips, we first selected a random 1000 covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.

```{r, eval = FALSE}
set.seed(42)
## Selecting 1000 random covariance matrices from the MCMCglmm
covar_matrices <- get.covar(combined_results, n = 100)
```

We will use these covariance matrices as the basis of the phylogenetic main axes of variation.
These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different levels: 
 * The whole bird phylogeny level ("phylogeny level");
 * And the named node levels ("clade levels").

These major axes where calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix).
In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigen values and the variance-covariance matrix's eigen vectors.

### Details for getting the major axes from the variance co-variance matrices

For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and of radius being composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

\begin{equation*}
O_{n} = 
    \begin{pmatrix}
        1 & 0 & \cdots & 0 \\
        0 & 1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & 1 \\
        -1 & 0 & \cdots & 0 \\
        0 & -1 & \cdots & 0 \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        0 & 0 & \cdots & -1 \\
    \end{pmatrix}
\end{equation*}

(in other words $O_{n}$ is the matrix representing each edge point of the unit hyper sphere of $n$ dimensions and of radius $1$);

And $O'_{n}$ the scaled matrix hyper sphere to the right confidence interval size using the $\chi^2$ distribution:

\begin{equation}
    O'_{n} = O_{n} \sqrt{\chi^2(0.95)}
\end{equation}

For the variance-covariance matrix $VCV_{n}$ of $n$ dimensions

\begin{equation*}
VCV_{n} = 
    \begin{pmatrix}
        \sigma(a) & \sigma(a,b) & \cdots & \sigma(a,n) \\
        \sigma(a,b) & \sigma(b) & \cdots & \sigma(b,n) \\
        \vdots  & \vdots  & \ddots & \vdots  \\
        \sigma(n,a) & \sigma(n,b) & \cdots & \sigma(n) \\
    \end{pmatrix}
\end{equation*}

and the eigenvectors **v** and the eigenvalues $\lambda$ satisfying the following eigen decomposition:

\begin{equation}
    VCV_{n} \bold{v} = \lambda \bold{v}
\end{equation}

We can get $M_{n}$ matrix containing all the edges' coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

\begin{equation}
    M_{n} = [(O'_{n}\sqrt{\lambda}) \times \bold{v}]^{\text{T}}
\end{equation}

Where $M_{1,n}$ is the major axes of the 0.95 CI hyper ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second second axes (minor axes in 2D), etc...

The detailed procedure was adapted from [李哲源's post on Stack Overflow](https://stackoverflow.com/questions/40300217/obtain-vertices-of-the-ellipse-on-an-ellipse-covariance-plot-created-by-care/40316331#40316331) and implemented in `beer::get.axis` ([specific procedure on lines 40-73](https://github.com/TGuillerme/elaboration_exploration_bird_beaks/blob/master/beer/R/get.axes.R)).

## References
